---
title: "Exercise 1"
author: "JMK CBRUG"
date: "2024-11-30"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

**Exercise 1: Accessing American Community Survey (ACS) Data with `tidycensus`**

**Objective:**

- Retrieve demographic data from the ACS.
- Perform basic data manipulation and visualization.

**Steps:**
1. **Install and load required packages:**
```{r package}
# Install packages if not already installed
#install.packages(c("tidycensus", "tidyverse"), repos = "http://cran.us.r-project.org")

# Load packages
library(tidycensus)
library(tidyverse)
```

2. **Set your API key for `tidycensus`:**
```{r api key}
# Set the API key (retrieved earlier)
#census_api_key(census_api_key, install = TRUE, overwrite = TRUE)
```

3. **Retrieve median household income data by state:**
```{r acs data}
income_data <- get_acs(
geography = "state",
variables = "B19013_001", # Median household income
year = 2023,
survey = "acs5")
```

4. **Inspect the data:**
```{r eda}
head(income_data)
```

5. **Create a bar chart of median household income by state:**
```{r}
income_data %>%
arrange(estimate) %>%
mutate(NAME = factor(NAME, levels = NAME)) %>%
ggplot(aes(x = NAME, y = estimate)) +
geom_col(fill = "steelblue") +
coord_flip() +
labs(
title = "Median Household Income by State (2023)",
x = "State",
y = "Median Household Income (USD)"
) +
theme_minimal()
```

**Notes:**
- Variable `B19013_001` corresponds to median household income.
- Data is retrieved for the year 2023 from the 5-year ACS estimates.

