---
title: "Exercise 3 PEP"
author: "JMK CBRUG"
date: "2024-11-30"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

**Exercise 3: Accessing Population Estimates Program (PEP) Data with `censusapi`**
**Objective:**
- Retrieve national population estimates.
- Analyze population trends over time.

**Steps:**
1. **Install and load the `censusapi` package:**
```{r packages}
# Install the package if not already installed
#install.packages("censusapi")

# Load the package
library(censusapi)

census_key <- "1628931ba3a0a3e59ffe62387248b2f340e3c1e2"
```

2. **Retrieve population estimates:**
```{r api}
apis <- listCensusApis() 

pep_apis <- apis[grepl("pep", 
                       apis$name, 
                       ignore.case = TRUE), ]

head(pep_apis)

pep_population_apis <- apis[grepl("pep/population", 
                                  apis$name, 
                                  ignore.case = TRUE), ]


# List the variables available

pep_variables <- listCensusMetadata(
    name = "2021/pep/population",
    type = "variables")

head(pep_variables)


# Define the dataset and variables you want to retrieve

dataset_pep <- "pep/population"

variables_pep <- c("REGION",      # Geographical region
                   "NPOPCHG_2020",  # Numeric Change in Population, April 1, 2020 Estimates Base to July 1, 2020
                   "NPOPCHG_2021",  # Numeric Change in Population, July 1, 2020 to July 1, 2021
                   "POP_2020",      # Population Estimate, July 1, 2020
                   "POP_2021")      # Population Estimate, July 1, 2021

pep_data_21 <- getCensus(
    key = key,    
    name = dataset_pep,     # Specify the API endpoint
    vintage = "2021",       # Specify the year
    vars = variables_pep,   # Specify the variables desired
    region = "state:*")     # Specify the Geography to get


variables_pep_19 <- c("REGION",  # Geographical region
                      "POP")       # Annual Estimates of the Resident Population: April 1, 2010 to July 1, 2019

pep_data_19 <- getCensus(
    key = key,    
    name = dataset_pep,        # Specify the API endpoint
    vintage = "2019",          # Specify the year
    vars = variables_pep_19,   # Specify the variables desired
    region = "state:*")  %>%   # Specify the Geography to get
    rename(POP_2019 = POP)

colnames(pep_data_19)

pep_data_18 <- getCensus(
    key = key,    
    name = dataset_pep,        # Specify the API endpoint
    vintage = "2018",          # Specify the year
    vars = variables_pep_19,   # Specify the variables desired
    region = "state:*") %>%   # Specify the Geography to get
  rename(POP_2018 = POP)

colnames(pep_data_18)

pep_data_17 <- getCensus(
    key = key,    
    name = dataset_pep,        # Specify the API endpoint
    vintage = "2017",          # Specify the year
    vars = variables_pep_19,   # Specify the variables desired
    region = "state:*")  %>%   # Specify the Geography to get
  rename(POP_2017 = POP)        


colnames(pep_data_17)

pep_data_16 <- getCensus(
    key = key,    
    name = dataset_pep,        # Specify the API endpoint
    vintage = "2016",          # Specify the year
    vars = variables_pep_19,   # Specify the variables desired
    region = "state:*")  %>%   # Specify the Geography to get
  rename(POP_2016 = POP)       


colnames(pep_data_16)

pep_data_15 <- getCensus(
    key = key,    
    name = dataset_pep,        # Specify the API endpoint
    vintage = "2015",          # Specify the year
    vars = variables_pep_19,   # Specify the variables desired
    region = "state:*") %>%    # Specify the Geography to get
  rename(POP_2015 = POP)      


colnames(pep_data_15)

```

3. **Clean and prepare the data:**
```{r}
pep_data_21 <- dplyr::arrange(pep_data_21, state)

pep_data_19 <- dplyr::arrange(pep_data_19, state) %>%
  select(-c(REGION)) %>%
  left_join(pep_data_18) %>%
  left_join(pep_data_17) %>%
  left_join(pep_data_16) %>%
  left_join(pep_data_15)

pep_data <- left_join(pep_data_21, pep_data_19, 
                      join_by(state == state), keep = FALSE)

```

```{r geography}
## Add Census Regions

# Create a data frame with state information
state_fips_regions <- tibble(
  state_fips = c(
    "01", "02", "04", "05", "06", "08", "09", "10", "11", "12",
    "13", "15", "16", "17", "18", "19", "20", "21", "22", "23",
    "24", "25", "26", "27", "28", "29", "30", "31", "32", "33",
    "34", "35", "36", "37", "38", "39", "40", "41", "42", "44",
    "45", "46", "47", "48", "49", "50", "51", "53", "54", "55", 
    "56", "72"
  ),
  state_name = c(
    "Alabama", "Alaska", "Arizona", "Arkansas", "California", "Colorado", "Connecticut",
    "Delaware", "District of Columbia", "Florida", "Georgia", "Hawaii", "Idaho", "Illinois",
    "Indiana", "Iowa", "Kansas", "Kentucky", "Louisiana", "Maine", "Maryland", "Massachusetts",
    "Michigan", "Minnesota", "Mississippi", "Missouri", "Montana", "Nebraska", "Nevada",
    "New Hampshire", "New Jersey", "New Mexico", "New York", "North Carolina", "North Dakota",
    "Ohio", "Oklahoma", "Oregon", "Pennsylvania", "Rhode Island", "South Carolina", "South Dakota",
    "Tennessee", "Texas", "Utah", "Vermont", "Virginia", "Washington", "West Virginia", "Wisconsin",
    "Wyoming", "Puerto Rico"
  ),
  state_abbr = c(
    "AL", "AK", "AZ", "AR", "CA", "CO", "CT",
    "DE", "DC", "FL", "GA", "HI", "ID", "IL",
    "IN", "IA", "KS", "KY", "LA", "ME", "MD", "MA",
    "MI", "MN", "MS", "MO", "MT", "NE", "NV",
    "NH", "NJ", "NM", "NY", "NC", "ND",
    "OH", "OK", "OR", "PA", "RI", "SC", "SD",
    "TN", "TX", "UT", "VT", "VA", "WA", "WV", "WI",
    "WY", "PR"
  )
)

# Define the Census regions
region_definitions <- list(
  Northeast = c("CT", "ME", "MA", "NH", "RI", "VT", "NJ", "NY", "PA"),
  Midwest = c("IL", "IN", "MI", "OH", "WI", "IA", "KS", "MN", "MO", "NE", "ND", "SD"),
  South = c("DE", "FL", "GA", "MD", "NC", "SC", "VA", "DC", "WV", "AL", "KY", "MS", "TN", "AR", "LA", "OK", "TX"),
  West = c("AZ", "CO", "ID", "MT", "NV", "NM", "UT", "WY", "AK", "CA", "HI", "OR", "WA"),
  Other = c("PR")
)

# Assign regions to each state
state_fips_regions <- state_fips_regions %>%
  mutate(region = case_when(
    state_abbr %in% region_definitions$Northeast ~ "Northeast",
    state_abbr %in% region_definitions$Midwest ~ "Midwest",
    state_abbr %in% region_definitions$South ~ "South",
    state_abbr %in% region_definitions$West ~ "West",
    state_abbr %in% region_definitions$Other ~ "Puerto Rico",
    TRUE ~ NA_character_
  ))

# View the result
head(state_fips_regions)

# Add to data

pep_data_region <- pep_data %>%
  left_join(state_fips_regions, by = join_by(state == state_fips)) %>%
  select(-c(state, REGION.x)) %>%
  select(state_name, state_abbr, REGION.y, region, 
         POP_2021, POP_2020, POP_2019, 
         POP_2018, POP_2017, POP_2016,
         POP_2015)

```
4. **Plot population estimates over time:**
```{r}
pep_wide <- pep_data_region %>%
  rename(`2021` = POP_2021,
         `2020` = POP_2020,
         `2019` = POP_2019,
         `2018` = POP_2018,
         `2017` = POP_2017,
         `2016` = POP_2016,
         `2015` = POP_2015) 

pep_long <- pep_wide %>%
  pivot_longer(!state_name:region,
    names_to = "year",
    values_to = "population"
  ) %>%
  mutate(year = as.integer(year))


pep_long %>%
ggplot(aes(x = year, y = population)) +
geom_point(color = "blue") +
labs(
title = "US Population Estimates Over Time",
x = "Year",
y = "Population"
) +
theme_minimal()
```
**Notes:**
- The `pep/population` endpoint provides population estimates.
- The `DATE_CODE` variable represents different time points.
