---
title: "Exercise 1"
author: "JMK CBRUG"
date: "2024-11-30"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

**Exercise 1: Accessing American Community Survey (ACS) Data with `tidycensus`**

**Objective:**

-   Retrieve demographic data from the ACS.
-   Perform basic data manipulation and visualization.

**Steps:** 1. **Install and load required packages:**

```{r package}
# Install packages if not already installed
#install.packages(c("tidycensus", "tidyverse", "censusapi", "patchwork"), repos = "http://cran.us.r-project.org")

# Load packages
library(tidycensus)
library(tidyverse)
library(censusapi)
library(patchwork)
```

2.  **Set your API key for `tidycensus`:**

```{r api key}
# Set the API key (retrieved earlier)
#census_api_key(census_api_key, install = TRUE, overwrite = TRUE)
```

3a. **Explore the available ACS endpoints:**

```{r endpoints}
apis <- listCensusApis()   # Get general information about available datasets

head(apis, n = 21) # Look at acs/acs5 in 2023
```

3b. **Explore the available ACS variables:**

```{r acs}
acs5_23_var <- listCensusMetadata(     # Get metadata about a specified API endpoint
  vintage = "2023",  
  name = "acs/acs5",
    type = "variables")

head(acs5_23_var, n = 25)

```

4.  **Retrieve population and median household income data by state:**

```{r acs data}
income_data <- get_acs(
geography = "state",
variables = c("B01001_001", # Total population
              "B19013_001"), # Median household income
year = 2023,
survey = "acs5") %>%
  mutate(var = case_when(variable == "B01001_001" ~ "population",
                         variable == "B19013_001" ~ "med hh income"))
```

5.  **Inspect the data:**

```{r eda}
head(income_data)

str(income_data)

options(scipen = 999)  # disable scientific notation

ggplot(income_data, aes(y = estimate)) +
  geom_boxplot() +
  facet_grid(rows = vars(var), scales = "free")

```

5.  **Create a bar chart of median household income by state:**

```{r}
income_data %>%
  filter(var == "med hh income") %>%
  arrange(estimate) %>%
  mutate(NAME = factor(NAME, levels = NAME)) %>%
ggplot(aes(x = NAME, y = estimate)) +
    geom_col(fill = "steelblue") +
  coord_flip() +
  labs(title = "Median Household Income by State (2023)",
        x = "State",
        y = "Median Household Income (USD)") +
  theme_minimal()
```

6.  **Create a bar chart of median household income by state, using color to highlight population:**

```{r}

pop_data <- income_data %>%
  filter(var == "population")


income_data %>%
  filter(var == "med hh income") %>%
  arrange(estimate) %>%
  mutate(NAME = factor(NAME, levels = NAME)) %>%
ggplot(aes(x = NAME, y = estimate, fill = pop_data$estimate)) +
    geom_col() +
  coord_flip() +
  labs(title = "Median Household Income by State (2023)",
        x = "State",
        y = "Median Household Income (USD)",
        fill = "Population") +
  theme_minimal()
```



**Notes:** 
- Variable `B01001_001` corresponds to the estimated state population in 2023. 
- Variable `B19013_001` corresponds to median household income. 
- Data is retrieved for the year 2023 from the 5-year ACS estimates.
